\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A Guide for International Students

### 2023-11-06 demystify.tex
updates

\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[margin=1.0in]{geometry}
\usepackage{soul}
%\usepackage[small,compact]{titlesec} %very powerful
\usepackage{tcolorbox}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=blue,
    urlcolor=blue,
}
\usepackage{booktabs}

\newtcolorbox{cbox}[1][]
{
left=1pt,right=1pt,top=1pt,bottom=1pt,
colback=gray!10,
boxrule=0.5pt
}

\newenvironment{commentbox}[1][]{
\small
    \begin{cbox}
    \textbf{#1} 
 }{
   \end{cbox}
}

@@

\section{Summary}\label{sec:summary}
Here we summarize the main points of the document. This allows you to quickly skim through the document and decide which parts you want to read in detail.
\begin{enumerate}
\item Should you apply to a PhD in CS in the US?
\begin{itemize}
    \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\S\ref{sec:should}).
\end{itemize}
\item How is your application evaluated?
\begin{itemize}
    \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically 3 faculty (\S\ref{sec:evalapps}).
    \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so contact and introduce yourself (\S\ref{sec:contact}).
\end{itemize}
\item Application Materials
\begin{itemize}
    \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, personal statement, and recommendation letters.
    \item LORs are important, but only if they are personalized and talk about your research ability (\S\ref{sec:lor}).
    \item GRE \emph{is not} required by most schools (\S\ref{sec:grades}).
    \item Grades are important, but only if your school is well-known (\S\ref{sec:grades}).
    \item Personal statement is very important. Write it in such a way that makes you \emph{stand out} (\S\ref{sec:research-statement} and \S\ref{sec:improve-your-chance})
    \item Getting an interview is typically a \emph{good sign} (no one interviews weak candidates).
\end{itemize}
\item What to do after getting admitted?
\begin{itemize}
    \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\S\ref{sec:accepted}).
\end{itemize}
\item Funding
\begin{itemize}
    \item TA and RA are two main sources of funding.  TA (teaching assistantship) is provided by the department to help profs. with classes (e.g., grading). RA (research assistantship) is provided by profs. to help them with their research (\S\ref{sec:funding}).
\end{itemize}
\item Choosing School and Professors
\begin{itemize}
    \item Many schools do not offer PhD studies in CS and many CS professors \emph{cannot} be adviser, i.e., cannot graduate PhD students  (\S\ref{sec:schoolsandprofs}).
    \item Contacting a prof. for research opportunities is recommended, but do it \emph{properly} (\S\ref{sec:contact}).
\end{itemize}
\item Miscs and FAQS
\begin{itemize}
    \item Increasing your admission chance by being unique and standing out (\S\ref{sec:improve-your-chance}).
    \item You can successfully apply to CS PhD even if you have non-STEM background (\S\ref{sec:non-stem}).
    \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\S\ref{sec:non-us-differences}).
    \item Your PhD costs about \$400K in total, but you do not pay for it (\S\ref{sec:ra-cost}).
    \item Despite some miserable stories on social media, many PhD students have good mentors, supportive labmates, healthy working environment ... and are happy (\S\ref{sec:happy}).
\end{itemize}
\end{enumerate}

@@

There are various guides on writing statement, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. So I will not talk too much about statements. In short, talk about your research goal and vision, and convince us that you can achieve it through your experience, e.g., published papers, or if you work on some projects by yourself, talk about it. Also, use the statement to talk about stuff that admission committee members might not know about, e.g., your Github project with 1K+ stars or  your regular contributions to well-known open-source projects. Finally, talk about what makes you different than the other 1000 applicants that we are reviewing (e.g., see \S\ref{sec:improve-your-chance} for increasing your admission chance).

@@

\begin{commentbox}[Thanh:]
Vietnamese universities typically offer specialized programs, such as the talented engineer program at HUST, that have highly competitive entrance exams and a limited number of available slots (e.g., 30 per year). However, these programs often set higher requirements for students, including more demanding tests and assignments, resulting in lower GPAs and overall rankings. For example, an 3.5 GPA students from such talented programs are typically much better than a 4.0 GPA students not in those programs.  Similarly, variations in GPA standards exist among different universities, with technical universities generally having lower GPAs than economical universities. These make gaining admission in the US difficult as US faculty are not familiar with these issues.
\tcblower
\textbf{Vu}: Vietnamese students and even faculty often lament how this competitive/low-grading system hurts Vietnamese students when applying abroad. One way to mitigate this is making these issues known to admission committee in your statement.  Universities with Vietnamese profs are probably aware of them, but in general your letter writers \emph{and} you can explicitly mention these in their letters and your statement.
\end{commentbox}

@@

Another very effective way is taking graduate level courses in the topics you are interested in (remember: you have those 2 years to explore around).  Professors teaching these \emph{special topics courses} and \emph{research seminars} typically are recruiting students---giving you even a higher chance. Do well in the class, answer questions, talk to the profs after classes, etc---being stand out.  Many professors, including myself, prefer taking in new students this way.  It gives both the professor and student more time, e.g., a whole semester, to work and evaluate the relationship before making any commitment (sounds a bit like a marriage!).  
You can also ask if you can do an independent study or research with a prof. This can be informal (no credit) and takes place during the summer or winter break.  For example, I work with multiple students, many of whom are undergrads, this way. Many will drop because they find they don't like my research, but some will stay and become my PhD students.

@@

There's a misconception that in the US it's all business, with professors as bosses who pay students for their work, and if you do something nice implies you expect something in return. However, the reality is quite the opposite. While people can be straightforward and direct, they are also friendly and informal. You can call your professor by their first name (\S\ref{sec:address}), respectfully disagree and argue with that (and can actually gain respect doing so), seek their help (even on personal matters), come to their houses for parties or gathering, and give them small thoughtful gifts that they put on their desks.  Many students and professors maintain lifelong relationships, staying in touch through cards, emails, and calls, even after their academic journey ends. These connections are genuine and meaningful.

>>>>

\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[margin=1.0in]{geometry}
\usepackage{soul}
%\usepackage[small,compact]{titlesec} %very powerful
\usepackage[most]{tcolorbox}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=blue,
    urlcolor=blue,
}
\usepackage{booktabs}

% \newtcolorbox{cbox}[1][]
% {
% left=1pt,right=1pt,top=1pt,bottom=1pt,
% colback=gray!10,
% boxrule=0.5pt
% }

\newtcolorbox{mybox}{
enhanced,
boxrule=0pt,frame hidden,
borderline west={2pt}{0pt}{green!75!black},
colback=green!10!white,
sharp corners
}

\newenvironment{commentbox}[1][]{
\small
    \begin{mybox}
    {\small \textbf{#1}}
 }{
   \end{mybox}
}

@@

\section{Summary}\label{sec:summary}

Here we summarize the main points of the document. This allows you to quickly skim through the document and decide which parts you want to read in detail.
\begin{enumerate}
\item Should you apply to a PhD in CS in the US?
\begin{itemize}
    \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\S\ref{sec:should}).
\end{itemize}
\item How is your application evaluated?
\begin{itemize}
    \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically 3 faculty (\S\ref{sec:evalapps}).
    \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so contact and introduce yourself (\S\ref{sec:contact}).
\end{itemize}
\item Application Materials
\begin{itemize}
    \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, personal statement, and recommendation letters.
    \item LORs are important, but only if they are personalized and talk about your research ability (\S\ref{sec:lor}).
    \item GRE \emph{is not} required by most schools (\S\ref{sec:grades}).
    \item Grades are important, but only if your school is well-known (\S\ref{sec:grades}).
    \item Personal statement is very important. Write it in such a way that makes you \emph{stand out} (\S\ref{sec:research-statement} and \S\ref{sec:improve-your-chance})
    \item Getting an interview is typically a \emph{good sign} (no one interviews weak candidates).
\end{itemize}
\item What to do after getting admitted?
\begin{itemize}
    \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\S\ref{sec:accepted}).
\end{itemize}
\item Funding
\begin{itemize}
    \item TA and RA are two main sources of funding.  TA (teaching assistantship) is provided by the department to help profs. with classes (e.g., grading). RA (research assistantship) is provided by profs. to help them with their research (\S\ref{sec:funding}).
\end{itemize}
\item Choosing School and Professors
\begin{itemize}
    \item Many schools do not offer PhD studies in CS and many CS professors \emph{cannot} be adviser, i.e., cannot graduate PhD students  (\S\ref{sec:schoolsandprofs}).
    \item Contacting a prof. for research opportunities is recommended, but do it \emph{properly} (\S\ref{sec:contact}).
\end{itemize}
\item Miscs and FAQS
\begin{itemize}
    \item Increasing your admission chance by being unique and standing out (\S\ref{sec:improve-your-chance}).
    \item You can successfully apply to CS PhD even if you have non-STEM background (\S\ref{sec:non-stem}).
    \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\S\ref{sec:non-us-differences}).
    \item Your PhD costs about \$400K in total, but you do not pay for it (\S\ref{sec:ra-cost}).
    \item Despite some miserable stories on social media, many PhD students have good mentors, supportive labmates, healthy working environment ... and are happy (\S\ref{sec:happy}).
\end{itemize}
\end{enumerate}

@@

There are various guides on writing statement, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. So I will not talk too much about statements. In short, talk about your research goal and vision, and convince us that you can achieve it through your experience, e.g., published papers, or if you work on some projects by yourself, talk about it. Also, use the statement to talk about things that admission committee members might not know about, e.g., your Github project with 1K+ stars or your regular contributions to well-known open-source projects. Finally, talk about what makes you different than the other 1000 applicants that we are considering (e.g., see \S\ref{sec:improve-your-chance} for increasing your admission chance).

@@

\begin{commentbox}[Thanh:]
Vietnamese universities typically offer specialized programs, such as the talented engineer program at HUST, that have highly competitive entrance exams and a limited number of available slots (e.g., 30 per year). However, these programs often set higher requirements for students, including more demanding tests and assignments, resulting in lower GPAs and overall rankings. For example, an 3.5 GPA students from such talented programs are typically much better than a 4.0 GPA students not in those programs.  Similarly, variations in GPA standards exist among different universities, with technical universities generally having lower GPAs than economical universities. These make gaining admission in the US difficult as US faculty are not familiar with these issues.
\tcblower
\textbf{Vu}: Vietnamese students and even faculty often lament how this grading system hurts Vietnamese students applying abroad. One way to mitigate this is making these issues known to admission committee in your statement.  Universities with Vietnamese profs are probably aware of them, but in general your letter writers \emph{and} you can explicitly mention these in their letters and your statement.
\end{commentbox}

@@

Another very effective way is taking graduate level courses in the topics you are interested in (remember: you have those 2 years to explore).  Professors teaching these \emph{special topics courses} and \emph{research seminars} typically are recruiting students---giving you even a higher chance. Do well in the class, answer questions, talk to the profs after classes, etc---being stand out.  Many professors, including myself, prefer taking in new students this way.  It gives both the professor and student more time, e.g., a whole semester, to work and evaluate the relationship before making any commitment (sounds a bit like a marriage!).  
You can also ask if you can do an independent study or research with a prof. This can be informal (no credit) and takes place during the summer or winter break.  For example, I work with multiple students, many of whom are undergrads, this way. Many will drop because they find they don't like my research, but some will stay and become my PhD students.

@@

There's a misconception that in the US it's all business, with professors as bosses who pay students for their work, and you doing something nice implying you expect something in return. However, the reality is quite the opposite. While people can be straightforward and direct, they are also friendly and informal. You can call your professor by their first name (\S\ref{sec:address}), respectfully disagree and argue with that (and can actually gain respect doing so), seek their help (even on personal matters), come to their houses for parties or gathering, and give them small thoughtful gifts that they put on their desks.  Many students and professors maintain lifelong relationships, staying in touch through cards, emails, and calls, even after their academic journey ends. These connections are genuine and meaningful.
